Refinements for open automata

### 2022-07-01 2021-Refinement/paper/refinementarticle.tex
transitivity 3eme item

Establishing equivalences and refinement relations between programs is an important mean for 
verifying the correctness of programs. By establishing that the
behaviours of a modified program simulate those of to the source one, simulaiton and bisimulation relations  formalise the desired  relation between a specification and an implementation, 
 two equivalent implementations, or a program and its optimised implementation.
This article discusses a notion of refinement between \emph{open automata}, which are symbolic
behavioural models for communicating systems. 
Open automata may have \emph{holes} modelling elements of their
context, and can be composed by instantiation of the holes. This allows for a compositional approach for
verification of their behaviour.

@@

Automata are convenient for specifying and verifying systems, but most automata definitions allow only the specifications of finite closed systems. Indeed those systems are the ones we can veify efficiently, but programming often consist in writing open systems that should be interfaced with others, and with potentially unbound behaviours (at least statically).We investigate in our works the possibility to define open symbolic systems, verify some of their properties, and use them to compose in a structured way more complex systems.

Refinement relations verify that and implementation satisfies some properties with respect to a specification. Refinement entails that one system can be considered as a more precise version of the specification, featuring all the specified behaviours with more concrete details. In this article we mostly rely on a simulation point of view, where all the specified behaviour must be followed by the refined system but additionla behaviours may exist. However we also ensure that a while scenario, made of several steps, of the specification can also be simulated by the refined system, which is a bit richer than the traditional simulation relation.

Open automata were defined as a way to provide a semantics for open parameterized hierarchical LTSs used in verification tools.
An open automaton \cite{henrio:01299562} is a classical automaton with variables and holes. Variables makes autoomata symbolic and allow them to encode infinite state systems. Holes enables the composition of automata: an automaton with a hole is an operator that takes another automaton as parameter and reacts to the actions it emits; the composed automaton is a more precise automaton where the behaviour of one ``parameter'' of the main automaton has been provided.

@@

Families of values, or equivalently maps will be noted \(\mset{i \mapsto x_i}{i \in I}\), \(\mset{i \gets x_i}{i \in I}\) or \(x_i^{i \in I}\). % TODO: introduire le fait que \exists c_j^{j \in J} défini J et {j \mapsto c_j}
%For instance \(\mpar{ax}^{x \in \setR}\) represents a scaling function, \(c^{i \in I}\) is a constant function over \(I\).
%They will be used depending on what is more convenient.
%For instance \(\mbrc{\alpha \mapsto 1, \beta \mapsto 2, \gamma \mapsto 3}\) has no simple generating expression and is better represented with the finite version of first notation.
The disjoint union on set is noted \(\uplus\)\footnote{\(\uplus\) notation either supposes that the sets are disjoint or rename conflicting objects depending on the context}. Disjoint union is also used on maps.
%There are several ways of ensuring a union is disjoint, we will indifferently either suppose sets are disjoint or rename conflicting object (useful for variables).
%The disjoint union of two maps \(\varphi: I \to X\) and \(\psi: J \to Y\) with \(I \cap J = \emptyset\) is noted \(\varphi \uplus \psi\) and has the following signature \(I \uplus J \to X \cup Y\).
In a formula, a quantifier followed by a finite set will be used as a shorthand for the quantification on every variable in the set:
\(\forall \mbrc{a_1, \dots, a_n}, \exists \mbrc{b_1, \dots, b_m}, P\) means \(\forall a_1, \dots, \forall a_n, \exists b_1, \dots, \exists b_m, P\).

@@

\item  Finally, we have to prove the following holds:
\begin{multline*}
\bigforall \mpar{s_1, s_3} \in S_1 \times S_3, \bigsymb{\forall} \mpar{t_1, t_3} = \mpar{\OTx{1}{}{1}{1}, \OTx{3}{}{3}{3}} \in \fOT{s_1} \times \fOT{s_3} \\ 
\bigsymb{\bigexists} 
\mpar{t_{1x} = \OTx{1}{x}{1x}{1x} \in \fOT{s_1}}^{x \in X},
 V_1 \uplus V_3\uplus \biguplus_{x\in X}\fvars{t_{1x}} \uplus \fvars{t_1} \uplus \fvars{t_3} \vdash \hspace{6cm}\\  R_{13}\mpar{s_1, s_3} \wedge 	g_1 \wedge g_3 \wedge \alpha_1 = \alpha_3 
%\wedge \bigwedge_\subbox{j \in J'_1 \cap J'_3} \beta_{1j} = \beta_{3j} 
\implies \hspace{6cm}\\
%\exists
%\OTx{1}{a}{3}{3}\in \fOT{s_1}. \\
\bigsymb{\bigwedge_{x\in X}}
\mpar{
\alpha_1=\alpha_{1 x}  \land J'_{1x}= J'_3 \land \bigwedge_\subbox{j \in J'_3 } \beta_{3j} = \beta_{1xj}\land g_{1 x}\land
 R_{13}\mpar{s'_{1 x}, s'_3}\psubst{\psi_{1 x} \uplus \psi_3}}
\end{multline*}
Firstly, according to the refinement relation between  $A_1$ and $A_2$ we have:
\begin{multline*}
\bigforall \mpar{s_1, s_2} \in S_1 \times S_2, \bigsymb{\forall} \mpar{t_1, t_2} = \mpar{\OTx{1}{}{1}{1}, \OTx{2}{}{2}{2}} \in \fOT{s_1} \times \fOT{s_2} \\ 
\bigsymb{\bigexists} 
\mpar{t_{1x} = \OTx{1}{x}{1x}{1x} \in \fOT{s_1}}^{x \in X},
 V_1 \uplus V_2 \uplus \biguplus_{x\in X}\fvars{t_{1x}} \uplus \fvars{t_1} \uplus \fvars{t_2} \vdash \hspace{6cm}\\  R_{12}\mpar{s_1, s_2} \wedge 	g_1 \wedge g_2 \wedge \alpha_1 = \alpha_2 
%\wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} 
\implies \hspace{6cm}\\
%\exists
%\OTx{1}{a}{3}{3}\in \fOT{s_1}. \\
\bigsymb{\bigwedge_{x\in X}}
\mpar{
\alpha_1=\alpha_{1 x}  \land J'_{1x}= J'_2 \land \bigwedge_\subbox{j \in J'_2 } \beta_{2j} = \beta_{1xj}\land g_{1 x}\land
 R_{12}\mpar{s'_{1 x}, s'_2}\psubst{\psi_{1 x} \uplus \psi_2}}
\end{multline*}
Yet according to the relation between   $A_2$ and $A_3$ we have:

>>>>

Establishing equivalences and refinement relations between programs is an important mean for 
verifying the correctness of programs. By establishing that the
behaviours of a modified program simulate those of to the source one,  and bisimulation relations  formalise the desired  relation between a specification and an implementation, 
 two equivalent implementations, or a program and its optimised implementation.
This article discusses a notion of refinement between \emph{open automata}, which are symbolic
behavioural models for communicating systems. 
Open automata may have \emph{holes} modelling elements of their
context, and can be composed by instantiation of the holes. This allows for a compositional approach for
verification of their behaviour.

@@

Automata are convenient for specifying and verifying systems, but most automata definitions allow only the specifications of finite closed systems. Indeed those systems are the ones we can verify efficiently, but programming often consist in writing open systems that should be interfaced with others, and with potentially unbound behaviours (at least statically).We investigate in our works the possibility to define open symbolic systems, verify some of their properties, and use them to compose in a structured way more complex systems.

Refinement relations verify that and implementation satisfies some properties with respect to a specification. Refinement entails that one system can be considered as a more precise version of the specification, featuring all the specified behaviours with more concrete details. In this article we mostly rely on a simulation point of view, where all the specified behaviour must be followed by the refined system but additional behaviours may exist. However we also ensure that a while scenario, made of several steps, of the specification can also be simulated by the refined system, which is a bit richer than the traditional simulation relation.

Open automata were defined as a way to provide a semantics for open parameterized hierarchical LTSs used in verification tools.
An open automaton \cite{henrio:01299562} is a classical automaton with variables and holes. Variables makes automata symbolic and allow them to encode infinite state systems. Holes enables the composition of automata: an automaton with a hole is an operator that takes another automaton as parameter and reacts to the actions it emits; the composed automaton is a more precise automaton where the behaviour of one ``parameter'' of the main automaton has been provided.

@@

Families of values, or equivalently maps will be noted \(x_i^{i \in I}\), \(\mset{i \mapsto x_i}{i \in I}\) or \(\mset{i \gets x_i}{i \in I}\). % TODO: introduire le fait que \exists c_j^{j \in J} défini J et {j \mapsto c_j}
%For instance \(\mpar{ax}^{x \in \setR}\) represents a scaling function, \(c^{i \in I}\) is a constant function over \(I\).
%They will be used depending on what is more convenient.
%For instance \(\mbrc{\alpha \mapsto 1, \beta \mapsto 2, \gamma \mapsto 3}\) has no simple generating expression and is better represented with the finite version of first notation.
The disjoint union on set is noted \(\uplus\)\footnote{\(\uplus\) notation either supposes that the sets are disjoint or rename conflicting objects depending on the context}. Disjoint union is also used on maps.
%There are several ways of ensuring a union is disjoint, we will indifferently either suppose sets are disjoint or rename conflicting object (useful for variables).
%The disjoint union of two maps \(\varphi: I \to X\) and \(\psi: J \to Y\) with \(I \cap J = \emptyset\) is noted \(\varphi \uplus \psi\) and has the following signature \(I \uplus J \to X \cup Y\).
In a formula, a quantifier followed by a finite set will be used as a shorthand for the quantification on every variable in the set:
\(\forall \mbrc{a_1, \dots, a_n}, \exists \mbrc{b_1, \dots, b_m}, P\) means \(\forall a_1, \dots, \forall a_n, \exists b_1, \dots, \exists b_m, P\).

@@

\item  Finally, we have to prove the following holds:
\begin{multline*}
\bigforall \mpar{s_1, s_3} \in S_1 \times S_3, \bigsymb{\forall} \mpar{t_1, t_3} = \mpar{\OTx{1}{}{1}{1}, \OTx{3}{}{3}{3}} \in \fOT{s_1} \times \fOT{s_3} \\ 
\bigsymb{\bigexists} 
\mpar{t_{1x} = \OTx{1}{x}{1x}{1x} \in \fOT{s_1}}^{x \in X},
 V_1 \uplus V_3\uplus \biguplus_{x\in X}\fvars{t_{1x}} \uplus \fvars{t_1} \uplus \fvars{t_3} \vdash \hspace{6cm}\\  R_{13}\mpar{s_1, s_3} \wedge 	g_1 \wedge g_3 \wedge \alpha_1 = \alpha_3 
%\wedge \bigwedge_\subbox{j \in J'_1 \cap J'_3} \beta_{1j} = \beta_{3j} 
\implies \hspace{6cm}\\
%\exists
%\OTx{1}{a}{3}{3}\in \fOT{s_1}. \\
\bigsymb{\bigwedge_{x\in X}}
\mpar{
\alpha_1=\alpha_{1 x}  \land J'_{1x}= J'_3 \land \bigwedge_\subbox{j \in J'_3 } \beta_{3j} = \beta_{1xj}\land g_{1 x}\land
 R_{13}\mpar{s'_{1 x}, s'_3}\psubst{\psi_{1 x} \uplus \psi_3}}
\end{multline*}

 Consider $\mpar{s_{1}, s_{3}} \in R_{13}$ then there is a set of states $(s_{2p})^{p\in P}$ of $A_2$ relating  $s_{1}$ and $s_{3}$:
\[R_{13}(s_1,s_3)=\bigvee_{{p\in P}
}\mpar{R_{12}\mpar{s_1,s_{2p}} \wedge R_{23}\mpar{s_{2p},s_3}} \]

Let's consider any $s_{2p} \in (s_{2p})^{p\in P}$. According to the refinement relation between  $A_1$ and $A_2$ and Definition \ref{def:preservation} we have:

\begin{multline*}
\bigforall \mpar{s_1, s_{2p}} \in S_1 \times S_2, \bigsymb{\forall} \mpar{t_1, t_{2p}} = \mpar{\OTx{1}{}{1}{1}, \OTx{2p}{}{2p}{2p}} \in \fOT{s_1} \times \fOT{s_{2p}} \\ 
\bigsymb{\bigexists} 
\mpar{t_{1x} = \OTx{1}{x}{1x}{1x} \in \fOT{s_1}}^{x \in X},
 V_1 \uplus V_2 \uplus \biguplus_{x\in X}\fvars{t_{1x}} \uplus \fvars{t_1} \uplus \fvars{t_{2p}} \vdash \hspace{6cm}\\  R_{12}\mpar{s_1, s_{2p}} \wedge 	g_1 \wedge g_{2p} \wedge \alpha_1 = \alpha_{2p} 
%\wedge \bigwedge_\subbox{j \in J'_1 \cap J'_{2p}} \beta_{1j} = \beta_{2pj} 
\implies \hspace{6cm}\\
%\exists
%\OTx{1}{a}{3}{3}\in \fOT{s_1}. \\
\bigsymb{\bigwedge_{x\in X}}
\mpar{
\alpha_1=\alpha_{1 x}  \land J'_{1x}= J'_{2p} \land \bigwedge_\subbox{j \in J'_{2p} } \beta_{2pj} = \beta_{1xj}\land g_{1 x}\land
 R_{12}\mpar{s'_{1 x}, s'_{2p}}\psubst{\psi_{1 x} \uplus \psi_{2p}}}
\end{multline*}

Furthermore, according to the refinement relation between  $A_2$ and $A_3$ and Definition \ref{Def:OA-Refinement} we have  for any open transition $t_{2px}$ in $T_2$ originating from $s_{2p}$ the following:
\begin{align*}
\OTx{2p}{x}{2px}{2px} \in \fOT{s_{2p}}
\end{align*}
there exists an indexed family of OTs originating from $s_3$: 
\begin{align*}
\mpar{\OTx{3}{py}{3py}{3py} \in \fOT{s_3}}^{y \in Y} 
\end{align*}
such that $\forall y \in Y, J'_{2p} \cap H = J'_{3py} \cap H$ and
\begin{multline*}
%\everymath{\displaystyle}
V_2 \uplus V_3 \uplus \fvars{t_{2p}} \vdash\\
R_{23}\mpar{s_{2p}, s_{3}} \wedge g_{2p} \implies  \operatorname*{\bigsymb{\bigvee}}_{y \in Y} \mpar{\everymath{\displaystyle}
\begin{array}{l}
			 \alpha_{2p}=\alpha_{3py} \wedge \bigwedge_\subbox{j \in J'_{3py} \cap H} \beta_{2pj}=\beta_{3pyj} \nwedge \\
			g_{3py} \wedge R_{23}\mpar{s'_{2p}, s'_{3y}}\psubst{\psi_{2p} \uplus \psi_{3py}}
		\end{array}}  
\end{multline*}

Based on             the above two properties, we get:
\begin{multline*}
R_{12}\mpar{s_1, s_{2p}} \wedge  R_{23}\mpar{s_{2p}, s_{3}} \wedge	g_1 \wedge g_{2p} \wedge \alpha_1 = \alpha_{2p} 
%\wedge \bigwedge_\subbox{j \in J'_1 \cap J'_{2p}} \beta_{1j} = \beta_{2pj} 
\implies \hspace{6cm}\\
%\exists
%\OTx{1}{a}{3}{3}\in \fOT{s_1}. \\
\bigsymb{\bigwedge_{x\in X}}
\operatorname*{\bigsymb{\bigvee}}_{y \in Y} \mpar{\mpar{\everymath{\displaystyle}
\begin{array}{l}
			 \alpha_{2p}=\alpha_{3py} \wedge \bigwedge_\subbox{j \in J'_{3py} \cap H} \beta_{2pj}=\beta_{3pyj} \nwedge \\
			g_{3py} \wedge R_{23}\mpar{s'_{2p}, s'_{3y}}\psubst{\psi_{2p} \uplus \psi_{3py}}
		\end{array}}\bigwedge
\mpar{\everymath{\displaystyle}\begin{array}{l}		
\alpha_1=\alpha_{1 x}  \land J'_{1x}= J'_{2p} \land \bigwedge_\subbox{j \in J'_{2p}} \beta_{2pj} = \beta_{1xj} \land\\ g_{1 x} \land
 R_{12}\mpar{s'_{1 x}, s'_{2p}}\psubst{\psi_{1 x} \uplus \psi_{2p}}
 \end{array}}
 }
\end{multline*}
By  rearranging the terms in the formula we get:
\begin{multline*}
R_{12}\mpar{s_1, s_{2p}} \wedge  R_{23}\mpar{s_{2p}, s_{3}} \wedge	g_1 \wedge g_{2p} \wedge \operatorname*{\bigsymb{\bigvee}}_{y \in Y}\mpar{ \alpha_1 = \alpha_{2p} \wedge \alpha_{2p}=\alpha_{3py} \wedge g_{3py} }
%\wedge \bigwedge_\subbox{j \in J'_1 \cap J'_{2p}} \beta_{1j} = \beta_{2pj} 
\implies \hspace{6cm}\\
%\exists
%\OTx{1}{a}{3}{3}\in \fOT{s_1}. \\
\bigsymb{\bigwedge_{x\in X}}
\operatorname*{\bigsymb{\bigvee}}_{y \in Y} 
\mpar{\everymath{\displaystyle}\begin{array}{l}		
\alpha_1=\alpha_{1 x}  \land J'_{1x}= J'_{2p} \land \bigwedge_\subbox{j \in J'_{2p}} \beta_{2pj} = \beta_{1xj} \land  \bigwedge_\subbox{j \in J'_{3py} \cap H} \beta_{2pj}=\beta_{3pyj} \\ g_{1 x}  \land
 R_{12}\mpar{s'_{1 x}, s'_{2p}} \wedge R_{23}\mpar{s'_{2p}, s'_{3y}} \psubst{\psi_{1 x} \uplus \psi_{2p}\uplus \psi_{3py}}
 \end{array}}
\end{multline*}

The formula is valid for all  $s_{2p} \in (s_{2p})^{p\in P}$. To build $R_{13}$ we need to rely on the disjunction of all possible path to relate $s_1$ and $s_3$:
\begin{multline*}
\operatorname*{\bigsymb{\bigvee}}_{p \in P} \mpar{R_{12}\mpar{s_1, s_{2p}} \wedge  R_{23}\mpar{s_{2p}, s_{3}} \wedge	g_1 \wedge g_{2p} \wedge \operatorname*{\bigsymb{\bigvee}}_{y \in Y}\mpar{ \alpha_1 = \alpha_{3py} \wedge g_{3py}}}
%\wedge \bigwedge_\subbox{j \in J'_1 \cap J'_{2p}} \beta_{1j} = \beta_{2pj} 
\implies \hspace{6cm}\\
%\exists
%\OTx{1}{a}{3}{3}\in \fOT{s_1}. \\
\bigsymb{\bigwedge_{x\in X}}
\operatorname*{\bigsymb{\bigvee}}_{p \in P}\operatorname*{\bigsymb{\bigvee}}_{y \in Y} 
\mpar{\everymath{\displaystyle}\begin{array}{l}		
\alpha_1=\alpha_{1 x}  \land J'_{1x}= J'_{2p} \land \bigwedge_\subbox{j \in J'_{2p}} \beta_{2pj} = \beta_{1xj} \land  \bigwedge_\subbox{j \in J'_{3py} \cap H} \beta_{2pj}=\beta_{3pyj} \\ g_{1 x}  \land
 R_{12}\mpar{s'_{1 x}, s'_{2p}} \wedge R_{23}\mpar{s'_{2p}, s'_{3y}} \psubst{\psi_{1 x} \uplus \psi_{2p}\uplus \psi_{3py}}
 \end{array}}
\end{multline*}

--------------------------------------

Firstly, according to the refinement relation between  $A_1$ and $A_2$ we have:
\begin{multline*}
\bigforall \mpar{s_1, s_2} \in S_1 \times S_2, \bigsymb{\forall} \mpar{t_1, t_2} = \mpar{\OTx{1}{}{1}{1}, \OTx{2}{}{2}{2}} \in \fOT{s_1} \times \fOT{s_2} \\ 
\bigsymb{\bigexists} 
\mpar{t_{1x} = \OTx{1}{x}{1x}{1x} \in \fOT{s_1}}^{x \in X},
 V_1 \uplus V_2 \uplus \biguplus_{x\in X}\fvars{t_{1x}} \uplus \fvars{t_1} \uplus \fvars{t_2} \vdash \hspace{6cm}\\  R_{12}\mpar{s_1, s_2} \wedge 	g_1 \wedge g_2 \wedge \alpha_1 = \alpha_2 
%\wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} 
\implies \hspace{6cm}\\
%\exists
%\OTx{1}{a}{3}{3}\in \fOT{s_1}. \\
\bigsymb{\bigwedge_{x\in X}}
\mpar{
\alpha_1=\alpha_{1 x}  \land J'_{1x}= J'_2 \land \bigwedge_\subbox{j \in J'_2 } \beta_{2j} = \beta_{1xj}\land g_{1 x}\land
 R_{12}\mpar{s'_{1 x}, s'_2}\psubst{\psi_{1 x} \uplus \psi_2}}
\end{multline*}
Yet according to the refinement  relation between   $A_2$ and $A_3$ and Definition \ref{Def:OA-Refinement} we have:
